title: "Google data analytics capstone: Cyclistic dataset"
author: "Lawrence Ea"
date: "2023-05-26"
output: 
  html_document: default
  pdf_document: default
  
#Background
Cyclistic is a bike-share initiative that stands out with its extensive fleet of 5,824 bicycles and 692 docking stations. 
What sets Cyclistic apart is its commitment to inclusivity by offering a variety of bike options, including reclining bikes, hand tricycles, and cargo bikes. 

This ensures that people with disabilities and those who cannot ride a standard two-wheeled bike can also participate in bike-sharing. 
While the majority of riders prefer traditional bikes, approximately 8% utilize the assistive options. While many users ride for leisure, a significant portion, around 30%, rely on Cyclistic bikes for their daily commute to work.


##Questions to solve
How does casual and members use cyclistic bikes differently?
How to convert casual riders to annual members?
How can cyclistic use digital media to influence casual rider's decision to become members?

#PREPARE
##I downloaded the 2019 Cyclistic dataset split in 4 Quarters




## Setting up my environment
```{r loading packages}
library(tidyverse)  #helps wrangle data
library(lubridate)  #wrangle date attributes
library(ggplot2)  #data visualization
library(readr) #Read csv files
library(dplyr) #For data manipulation
```

#PROCESS (Loading and cleaning the dataset)
```{r include=FALSE}
Q1 <- read_csv("Capstone 2019 Quarterly/Divvy_Trips_2019_Q1.csv")

Q2 <- read_csv("Capstone 2019 Quarterly/Divvy_Trips_2019_Q2.csv")

Q3 <- read_csv("Capstone 2019 Quarterly/Divvy_Trips_2019_Q3.csv")

Q4 <- read_csv("Capstone 2019 Quarterly/Divvy_Trips_2019_Q4.csv")

#NAMING COLUMN NAMES CONSISTENTLY TO APPEND TABLES PROPERLY

(Q4 <- rename(Q4
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(Q3 <- rename(Q3
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(Q2 <- rename(Q2
              ,ride_id = trip_id
              ,rideable_type = bikeid 
              ,started_at = start_time  
              ,ended_at = end_time  
              ,start_station_name = from_station_name 
              ,start_station_id = from_station_id 
              ,end_station_name = to_station_name 
              ,end_station_id = to_station_id 
              ,member_casual = usertype))

(Q1 <- rename(Q1
              ,ride_id = trip_id
  ,rideable_type = bikeid 
,started_at = start_time  
,ended_at = end_time  
,start_station_name = from_station_name 
,start_station_id = from_station_id 
,end_station_name = to_station_name 
,end_station_id = to_station_id 
,member_casual = usertype))

# CONVERT RIDE ID AND RIDEABLE TYPE TO CHARACTER


Q4 <-  mutate(Q4, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
Q3 <-  mutate(Q3, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
Q2 <-  mutate(Q2, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
Q1 <-  mutate(Q1, ride_id = as.character(ride_id)
              ,rideable_type = as.character(rideable_type)) 




#Correct code: combined mutate with POSIXct function
Q2 <- Q2 %>%
  mutate(started_at = as.POSIXct(started_at, format = "%m/%d/%Y %H:%M"))
Q2 <- Q2 %>%
  mutate(ended_at = as.POSIXct(ended_at, format = "%m/%d/%Y %H:%M"))

data_2019 <- bind_rows(Q1,Q2,Q3,Q4)






# Remove columns : lat, long, birthyear, and gender
all_trips_2019 <- data_2019 %>%  
  select(-c(birthyear, gender,))

##Data CLeaning step
colnames(all_trips_2019)  #column names
nrow(all_trips_2019)  #Number of rows in data frame?
dim(all_trips_2019)  #Rows and columns count
head(all_trips_2019)  #first 6 rows of data frame.  Also tail(all_trips)
str(all_trips_2019)  #list of columns and data types (numeric, character, etc)

#Statistical summary
#summary(all_trips_v2) 

#member and casual count
table(all_trips_2019$member_casual)

all_trips_2019 <-  all_trips_2019 %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))

# Added columns fordate, month, day, and year of each ride
all_trips_2019$date <- as.Date(all_trips_2019$started_at) #The default format is yyyy-mm-dd
all_trips_2019$month <- format(as.Date(all_trips_2019$date), "%m")
all_trips_2019$day <- format(as.Date(all_trips_2019$date), "%d")
all_trips_2019$year <- format(as.Date(all_trips_2019$date), "%Y")
all_trips_2019$day_of_week <- format(as.Date(all_trips_2019$date), "%A")

all_trips_2019$ride_length <- difftime(all_trips_2019$ended_at,all_trips_2019$started_at)


#Remove Bad data (remove all trips with 0 or negative duration)

all_trips_v2 <- all_trips_2019[!(all_trips_2019$start_station_name == "HQ QR" | all_trips_2019$ride_length<0),]
#I removed trips with Null values
all_trips_nonull <- na.omit(all_trips_v2) 


#Convert tripduration from factor to Numeric
is.factor(all_trips_v2$tripduration)
all_trips_v2$tripduration <- as.numeric(as.character(all_trips_v2$tripduration))
is.numeric(all_trips_v2$tripduration)

# ANALYZE (Summary of analysis)

###
Total rides: 3,119,296

Average ride length: 24.11 minutes

Busiest time for members and casual: 5pm

Busiest month: August

Busiest Day: Wednesday

###Comparing Member vs. Casual

#Average ride length (It is evident based on the average ride length that members and casual use the bike-sharing service for different purpose)
  *Member: 25.31 minutes
  *Casual: 106.97 minutes
  
#Total rides
  *Member: 2,411,956 (77.32%)
  *Casual:  707,340 (22.68%)

#Busiest day
  *Member: Wednesday
  *Casual: Saturday
  
#Busiest month
  *Member: August
  *Casual: August
  
#Ride legnth comparison
  *Member: consistent users thru the weekdays
  *Casual: Higher demand on weekends


### Analysis on ride length (results)
#Mean =1447.354 seconds
Median = 706 seconds
min = 61 seconds
max = 10628400 (Seconds), 177,140 minutes

#The max length is a possible outlier because of the duration of bike rental


mean(all_trips_2019$tripduration) #straight average (total ride length / rides)
median(all_trips_2019$tripduration) #midpoint number in the ascending array of ride lengths
max(all_trips_2019$tripduration) #longest ride
min(all_trips_2019$tripduration) #shortest ride

# Average ride in a week
aggregate(all_trips_v2$tripduration ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

#Arrange weekday
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))


# analyze ridership data by type and day of the week
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by casual/member
  summarise(tripduration = n()		#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```



#SHARE (Data Visualizations)
### These are the visualizations to further explain the findings on Cyclistic dataset

#### Total number of trips by customer in year 2019
##### Here we will have an overview about number of trips taken by usertype, almost 2.5 million of users are members while approximately 750,000 users are casual. This suggest that majority of customer base consist of members who are likely recurring users of the bike sharing service.

```{r echo=FALSE}
all_trips_nonull %>%
  group_by(member_casual) %>%
  summarise(total_rows = n()) %>%
  ggplot(aes(x=member_casual, y=total_rows, fill=member_casual)) +
  labs(title = "Total trips by Customer") +
  labs(x="User type", y="Total rides") +
  theme(axis.text.y=element_text(angle = 45)) +
  geom_col(stat="identity") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE), 
                     breaks = c(0, 0.5e6, 1e6, 1.5e6, 2e6, 2.5e6))
```


#### Number of rides by usertype in a week
##### Here we will see the total number of rides taken by customer distributed in days of week. It is evident in the visualization that members consistently take more rides than casual users throughout the week.

##### This indicates that members primarily use the service for commuting or as a regular transportation. On the other hand, Casual riders utilize the service on weekends which suggests that they use it for leisure and occasional purposes.
```{r echo=FALSE}
all_trips_nonull %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x = "Day of week", y = "Number of rides") +
  scale_y_continuous(labels = scales::number_format(scale = 1, accuracy = 1))
```


#### Average length of bike rides by user type
##### Here we can compare the average length of use by member and casual riders
##### It is evident in the visualization that Members use the bikes for less than 50 minutes which translates to transportation going to work or school, While casual users on the other hand has longer duration due to leisure purpose or exploring the city throughout the day.
```{r echo=FALSE}
all_trips_nonull %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(x="Day of week", y="Ride duration in minutes")
```

#### Hourly demand of bike in a day
##### For members there are 2 noticeable peak in demand: one in the morning at 8 and another in the afternoon at 5pm. These hours align with typical commuting hours, which supports our previous conclusion that members rely on this bike sharing service for their daily transportation.

##### The bike demand for casual riders started to pick up at around 9 am and peak at 5pm which is identical with the peak demand for members.
```{r echo=FALSE}
all_trips_nonull %>%
  mutate(hour = hour(started_at)) %>%
  group_by(hour, member_casual) %>%
  summarise(ride_count = n()) %>%
  ggplot(aes(x = hour, y = ride_count, color = member_casual)) +
  labs(title = "Hourly Demand of bike for 2019", subtitle="Rides by customer type") +
  scale_x_continuous(breaks = seq(0, 23, 1)) +
  scale_y_continuous(breaks = c(50000, 100000, 150000, 200000, 225000, 250000, 275000), labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  geom_line() +
  theme_minimal() + 
  labs(x="Month", y="Number of rides", )
```

#### Total rides per month in a year
##### Here we will see the busiest month which is summer and early fall of 2019. We can see that total number of rides for members are more than double the rides per month of casual riders.
```{r echo=FALSE}
all_trips_nonull %>%
  group_by(month, member_casual) %>%
  summarize(total_rides = n()) %>% 
  ggplot(aes(x = month, y = total_rides, fill=member_casual)) +
  geom_col(position = position_dodge()) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  labs(x = "Month", y = "Total Rides", fill = "Member Type", title = "Total Rides per Month in 2019") +
  theme(legend.position = "bottom")

```
##Findings:
-Almost 2.5 million of users are members while approximately 750,000 users are casual. This suggest that majority of customer base consist of members who are likely recurring users of the bike sharing service.

-Members took more than 70% of the rides for 2019. On the other hand, while casual riders has longer ride length average throughout the week.

-This suggest that casual riders use the service for leisure and recreation.

-Members utilize the bikeshare service for daily transportation

-Busiest days for members are weekdays while on the weekends casual members exceed the 100k mark on number of rides.

-The bike demand for casual riders started to pick up at around 9 am and peak at 5pm which is identical with the peak demand for members.


#ACT (Recommendations)
-During the period from July to October, there is a significant number of casual riders. Analysis of the data reveals that on weekdays, the number of casual riders ranges from 80,000 to 90,000 on a single day. This presents a potential impact on profits if these riders were to become members.

-Consider monthly membership option with more flexible pricing would be ideal, to cater to the needs of locals, tourists, and business travelers.

-Regular weekday casual users who use the service between July and October should be encouraged to consider signing up for a monthly membership.

-Utilize social media to promote activities and convenience of the bike-sharing service, Engaging content showcasing exciting bike-related activities and highlighting the ease of returning bikes to different docking stations across the city would be effective.

-Emphasize the environmental(carbon emissions), Financial(Cheaper than driving) and Health benefits, Cyclistic can encourage local commuters who currently drive to work to switch to biking during the summer months. This can be achieved by offering attractive incentives through the monthly pass option.

-Considering monthly passes can solve the hurdles of covnerting casual riders to members.
